---
title: "Module 10"
author: "Cat Metcalf"
date: "2023-09-30"
output: html_document
---

# Module 10 Classical Hypothesis Testing

Classical or frequentist hypothesis testing (a.k.a. parametric statistics) involves formally stating a claim - the **null hypothesis** - which is then followed up by statistical evaluation of the null versus an alternative hypotheses. The null hypothesis is interpreted as a baseline hypothesis and is the claim that is assumed to be true. The **alternative hypothesis** is the conjecture that we are testing.

H0  = null hypothesis = a sample statistic shows no deviation from what is expected or neutral

HA = alternative hypothesis = a sample statistic deviates more than expected by chance from what is expected or neutral

HA > H0, which constitutes an “upper one-tailed test (i.e., our sample statistic is greater than that expected under the null)

HA < H0, which constitutes an “lower one-tailed test (i.e., our sample statistic is less than that expected under the null)

HA ≠ H0, which constitutes an “two-tailed test (i.e., our sample statistic is different, maybe greater maybe less, than that expected under the null)

There are then four possible outcomes to our statistical decision:

What is True	What We Decide	Result
H0	H0	Correctly ‘accept’ the null
H0	HA	Falsely reject the null (Type I error)
HA	H0	Falsely ‘accept’ the null (Type II error)
HA	HA	Correctly reject the null

**p value** = the probability of our obtaining a test statistic that is as high or higher than our calculated one by chance, assuming the null hypothesis is true.

The particular value of our test statistic is determined by both the difference between the original sample statistic and the expected null value (e.g., the difference between the mean of our sample and the expected population mean) and the standard error of the sample statistic. The value of the test statistic (i.e., the distance of the test statistic from zero) and the shape of the null distribution are the sole drivers of the smallness of the p value. The p value effectively represents the area under the sampling distribution associated with test statistic values as or more extreme than the one we observed.

How do we calculate the p value?

Specify a test statistic (e.g., the mean)

Specify our null distribution

Calculate the tail probability, i.e., the probability of obtaining a statistic (e.g., a mean) as or more extreme than was observed assuming that null distribution

***

Let’s do an example where we try to evaluate whether the mean of a single set of observations is significantly different than expected under a null hypothesis… i.e., this is a ONE-SAMPLE test.

Suppose we have a vector describing the adult weights of vervet monkeys trapped in South Africa during the 2016 trapping season. We have the sense they are heavier than vervets we trapped in previous years, which averaged 4.9 kilograms. The mean is 5.324 kilograms. Is the mean significantly greater than our expectation?

```{r}
library(curl)
```

```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/vervet-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
mean(d$weight)
```

What is our H0? vervets in previous years were 4.9 kilograms
What is our HA? vervets trapped recently are heavier than the previous years
What is the hypothesis we want to test? Is it two-tailed? Upper-tailed? Lower-tailed? The HA, and it is upper-tailed.
Calculate the mean, standard deviation, and SEM from our sample
```{r}
mu <- 4.9
x <- d$weight
m <- mean(x)
s <- sd(x)
n <- length(x)
sem <- s/sqrt(n)
```

```{r}
z <- (m - mu)/sem
z
```
z is a quantile, estimated number of standard errors of the mean away from the population mean that the sample falls

We want the probability of seeing a z this large or larger by chance.
```{r}
p <- 1 - pnorm(z)
p
```

```{r}
p <- pnorm(z, lower.tail = FALSE)
p
```

Use the T distribution
```{r}
p <- 1 - pt(z, df = n - 1)
p
```
```{r}
p <- pt(z, df = n - 1, lower.tail = FALSE)
p
```

R has built into it a single function, t.test(), that lets us do all this in one line. We give it our data and the expected population mean, μ, along with the kind of test we want to do
```{r}
t <- t.test(x = x, mu = mu, alternative = "greater")
t
```

```{r}
lower <- m - qt(1 - 0.05/2, df = n - 1) * sem
upper <- m + qt(1 - 0.05/2, df = n - 1) * sem
ci <- c(lower, upper)
ci  # by hand
```

```{r}
t <- t.test(x = x, mu = mu, alternative = "two.sided")
ci <- t$conf.int
ci  # using t test
```

So our conclusion, then, would be to reject the H0 that the weights of the sample of monkeys from the 2016 trapping season come from the same distribution as the weights of monkeys from previous trapping seasons based on the average, since the average from previous seasons falls outside the 95% confidence interval for the t distribution based on the sample average from the 2016 trapping season.

In other words, the 2016 trapping season has vervet monkeys whose weights are significantly heavier than the average from previous trapping seasons (p < 0.01).

## Challenge 1
Adult lowland woolly monkeys are reported to have an average body weight of 7.2 kilograms. You are working with an isolated population of woolly monkeys from the Peruvian Andes that you think may be a different species from the lowland form, and you collect a sample of 15 weights from adult individuals at that site. From your sample, you calculate a mean of 6.43 kilograms and a standard deviation of 0.98 kilograms.

Perform a hypothesis test of whether body weights in your population are different from the reported average for lowland woolly monkeys by setting up a “two-tailed” hypothesis, carrying out the analysis, and interpreting the p value (assume the significance level is α=0.05). Your sample is < 30, so you should use the t distribution and do a t test. Do your calculations both by hand and using the t.test() function and confirm that they match.

```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall21/woolly-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
x <- d$weight
m <- mean(x)
s <- sd(x)
n <- length(x)
sem <- s/sqrt(n)
mu <- 7.2
t <- (m - mu)/sem
t
```

```{r}
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
test <- t < -crit || t > crit  # boolean test as to whether t is larger than the critical value at either tail
test <- abs(t) > crit
t.test(x = x, mu = mu, alternative = "two.sided")
```
Based on our conclusions for the vervet sample, what do the results of this comparison across woolly monkey populations mean? 

Sometimes we want to compare two groups of measurements to one another, which boils down to a hypothesis test for the difference between two means, μ1 and μ2. The null hypothesis is that the difference between the means is zero.

Before getting to the appropriate test, there are a couple of things that we need to consider:

[1] How, if at all, are the two samples related to one another? Sometimes we may have PAIRED samples (e.g., the same individuals before and after some treatment) and sometimes the samples are UNPAIRED or INDEPENDENT (e.g., weights for different samples of black-and-white colobus monkeys collected in the rainy versus dry seasons).

[2] Are the variances in the two samples roughly equal or not? E.g., if we are comparing male and female heights, are the variances comparable?

Samples with Unequal Variances
For the most generic case, where the two samples are independent and we cannot assume the variances of the two samples are equal, we can do what is called Welch’s t test where our test statistic is:

T=(review module text)

where:

 and  = means of observations in each sample group
μ = expected difference in means between sample groups under the null hypothesis, which is usually zero
s1 and s2 = standard deviations of each sample group
n1 and n2 = numbers of observations in each sample group

## Challenge 2
Let’s do an example. Load in a file of black-and-white colobus weights and examine the str() of the file. Then, create 2 vectors, x and y, for male and female weights. Plot these in boxplots side by side and then calculate the mean, sd, and sample size for both males and females.
```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/colobus-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
x <- d$weight[d$sex == "male"]
y <- d$weight[d$sex == "female"]
par(mfrow = c(1, 2))
boxplot(x, ylim = c(4.5, 8), main = "Weight (kg)", xlab = "Males")
boxplot(y, ylim = c(4.5, 8), main = "Weight (kg)", xlab = "Females")
```

```{r}
m1 <- mean(x)
m2 <- mean(y)
mu <- 0  # you could leave this out... the default argument value is 0
s1 <- sd(x)
s2 <- sd(y)
n1 <- length(x)
n2 <- length(y)
```

Now calculate the t statistic and test the two-tailed hypothesis that the sample means differ.
```{r}
t <- (m2 - m1 - mu)/sqrt(s2^2/n2 + s1^2/n1)
t
```

```{r}
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
crit
```

```{r}
test <- t < -crit || t > crit  # boolean test
test <- abs(t) > crit
test
```

Number of degrees freedom:
```{r}
df <- (s2^2/n2 + s1^2/n1)^2/((s2^2/n2)^2/(n2 - 1) + (s1^2/n1)^2/(n1 - 1))
df
```

```{r}
t <- t.test(x = x, y = y, mu = 0, alternative = "two.sided")
t
```
How do we interpret these results?

There’s a simpler t statistic we can use if the variances of the two samples are more or less equal.

```{r}
s <- sqrt((((n1 - 1) * s1^2) + ((n2 - 1) * s2^2))/(n1 + n2 - 2))
t <- (m2 - m1 - mu)/(sqrt(s^2 * (1/n1 + 1/n2)))
t
```

```{r}
df <- n1 + n2 - 2
df
```

```{r}
t <- t.test(x = x, y = y, mu = 0, var.equal = TRUE, alternative = "two.sided")
t
```
A crude test for equality of variances is to divide the larger by the smaller and if the result is < 2, you can go ahead and used the pooled variance version of the test (which has many fewer degrees of freedom).

In our case, we cannot, since the ratio of variances exceeds 2…
```{r}
var(x)/var(y)
```

We can use the var.test() function to conduct an actual statistical test on the ratio of variances, which compares the ratio test statistic we just calculated to an F distribution. The F distribution is often used to model ratios of random variables and thus is useful in regression applications and, as here, for testing whether variances from two samples are different. It is dependent upon the specification of a pair of degrees of freedom values supplied as the arguments df1= and df2= (or inferred from the number of observations in each sample).

Below, the results of var.test() are saved to a variable. Calling the variable provides a brief descriptive summary.
```{r}
vt <- var.test(x, y)
vt
```

***
For a paired sample test, the null hypothesis is that the mean of individual paired differences between the two samples (e.g., before and after) is zero.

## Challenge 3
Let’s play with a sample… IQs of individuals taking a certain statistics course before and after a lecture on significance testing. Load in the iqs.csv data file, look at it, plot a barchart of values before and after and construct a paired t test to evaluate the means before and after.
```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/iqs.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r}
x <- d$IQ.before - d$IQ.after
m <- mean(x)
mu <- 0  # can leave this out
s <- sd(x)
n <- length(x)
sem <- s/sqrt(n)
par(mfrow = c(1, 2))
boxplot(d$IQ.before, ylim = c(115, 145), main = "IQ", xlab = "Before")
boxplot(d$IQ.after, ylim = c(115, 145), main = "IQ", xlab = "After")
```

```{r}
t <- (m - mu)/sem
t
```

```{r}
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
crit
```

```{r}
test <- t < -crit || t > crit  # boolean test
test
```

```{r}
t.test(x, df = n - 1, alternative = "two.sided")
```

As we have seen, the sampling distribution of sample means for independent and identically distributed random variables is roughly normal (and, as shown by the CLT, this distribution increasingly approaches normal as sample size increases). Similarly, the sampling distribution for another kind of sample statistic, the number of “successes” x out of a series of k trials is also roughly normally distributed. If the true population proportion of “successes” is π, then the sampling distribution for the proportion of successes in a sample of size n is expected to be roughly normally distributed with mean = π and standard error = sqrt(π(1−π)/n).

Let’s set up a simulation to show this…

First we create a population of 500 “1”s and 500 “0”s, i.e., where π = 0.5…
```{r}
pop <- c(rep(0, 500), rep(1, 500))
```
Now, we will take 1000 random samples of size n=10 from that population and calculate the proportion of “1”s in each sample…
```{r}
pi <- 0.5
x <- NULL
n <- 10
for (i in 1:1000) {
    x[i] <- mean(sample(pop, size = n, replace = FALSE))  # taking the mean of a bunch of 0s and 1s yields the proportion of 1s!
}
m <- mean(x)
m
```

```{r}
s <- sd(x)
s
```

```{r}
pop_se <- sqrt(pi * (1 - pi)/n)
pop_se  # the se is an estimate of the sd of the sampling distribution
```

The same is true if we create a population of 800 “1”s and 200 “0”s, i.e., where π = 0.8…
```{r}
pop <- c(rep(0, 800), rep(1, 200))
pi <- 0.8
x <- NULL
n <- 10
for (i in 1:1000) {
    x[i] <- mean(sample(pop, size = n, replace = FALSE))  # taking the mean of a bunch of 0s and 1s yields the proportion of 1s!
}
m <- mean(x)
m
```

```{r}
s <- sd(x)
s
```

```{r}
pop_se <- sqrt(pi * (1 - pi)/n)
pop_se  # the se is an estimate of the sd of the sampling distribution
```

This normal approximation is true as long as n is fairly large and π is not close to 0 or 1. One rule of thumb is to check that both n×π and n×(1−π) are greater than 5.

With all this in mind, we can construct Z statistics for proportions like we constructed Z and T statistics for means and test those proportions for differences from an expected value or for differences between two sample proportions. The Z statistic for proportions takes the same general form as that for means…

## Challenge 4
A neotropical ornithologist working in the western Amazon deploys 30 mist nets in a 100 hectare (ha) grid. She monitors the nets on one morning and records whether or not a she captures any birds in the net (i.e., a “success” or “failure” for every net during a netting session). The following vector summarizes her netting results:
```{r}
v <- c(0, 1, 1, 1, 0, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0, 1, 1, 0, 0, 1, 1, 1, 0,
    1, 1, 0, 1, 0, 1, 1)
```
Her netting success over the previous three seasons suggests that she should catch birds in 80% of her nets. This season, she feels, her success rate is lower than in previous years. Does her trapping data support this hypothesis?

What is H0?
What is HA?
Are both n×π and n×(1−π) > 5?
Calculate z and the p value associated with z
Calculate the 95% CI around 

```{r}
phat <- mean(v)
phat
```

```{r}
pi <- 0.8
n <- 30
z <- (phat - pi)/sqrt(pi * (1 - pi)/30)
z
```

```{r}
p <- pnorm(z, lower.tail = TRUE)
p
```

We use the lower.tail=TRUE argument because we’re testing a lower-tailed one-tailed hypothesis. The 95% confidence interval can be estimated, based on the normal distribution, as follows:
```{r}
lower <- phat - qnorm(0.975) * sqrt(phat * (1 - phat)/30)
upper <- phat + qnorm(0.975) * sqrt(phat * (1 - phat)/30)
ci <- c(lower, upper)
ci
```

This approach using quantiles of the standard normal distribution is but one method of calculating CIs for proportion data, and is the CI referred to as a Wald confidence interval. Note that the CI does not include the value of π… rather, π > is greater than the upper bound of the CI, suggesting that the observed success rate is indeed lower than in previous years.

We can do the same test with a one-liner in R…
```{r}
pt <- prop.test(x = sum(v), n = length(v), p = 0.8, conf.level = 0.95, correct = FALSE,
    alternative = "less")
pt
```

Note that the CI is different than we calculated based on the normal distribution… prop.test() implements a slightly different procedure for estimating the CI rather than basing this on the normal distribution and the CLT.

How should she interpret these results? Was she right that her trapping results are off this year?

## Challenge 5
A biologist studying two species of tropical bats captures females of both species in a mist net over the course of week of nightly netting. For each species, the researcher records whether females are lactating or not. The two vectors below summarize the data for each species.
```{r}
v1 <- c(1, 0, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0,
    1, 0)
v2 <- c(1, 1, 0, 0, 1, 1, 1, 1, 0, 1, 0, 1, 1, 1, 0, 1, 0, 1, 1, 1, 1, 1, 0,
    0, 1, 1, 0, 1, 1, 1)
```

Based on your mist netting data, do the species differ significantly in the proportion of lactating females during this trapping season? What are H0 and HA?
```{r}
pstar <- (sum(v1) + sum(v2))/(length(v1) + length(v2))
pstar
```

```{r}
phat1 <- mean(v1)
phat1
```

```{r}
phat2 <- mean(v2)
phat2
```

```{r}
pi <- 0
z <- (phat2 - phat1)/sqrt((pstar * (1 - pstar)) * (1/length(v1) + 1/length(v2)))
z
```

```{r}
p <- 1 - pnorm(z, lower.tail = TRUE) + pnorm(z, lower.tail = FALSE)
p
```

```{r}
crit <- qnorm(1 - alpha/2)  # identify critical values
crit
```

```{r}
test <- p < -crit || p > crit  # boolean test
test
```

```{r}
pt <- prop.test(x = c(sum(v2), sum(v1)), n = c(length(v2), length(v1)), alternative = "two.sided",
    correct = FALSE)
pt
```
What are her results? Are the proportions of lactating females different across species during this trapping season?

**Summary of Z and T Tests**
Z and T tests are used to evaluate whether a given sample statistic (e.g., a mean or proportion) deviates significantly from what is expected under a null model or whether two samples statistics deviate significantly from one another.

*Module 10 Complete*