---
title: "Module 9"
author: "Cat Metcalf"
date: "2023-09-30"
output: html_document
---

# Module 9 Introduction to Statistical Inference

```{r}
library(curl)
```

The **critical value** is derived from the standardized version of a sampling distribution (e.g., the normal distribution) that corresponds the quantile limits we are interested in. For example, for the 95% CI around the mean, the critical value corresponds the range of quantiles above and below which we expect to see only 5% of the distribution of statistic values. This is equivalent to the ± 1 - (α/2) quantiles, where α=0.05, i.e., the ± 0.975 quantile that we have used before for calculating 95% CIs.

The **standard error** is the standard deviation of the sampling distribution, which, as noted above, is often estimated from the sample itself as σ/sqrt(n) but can also be calculated directly from the population standard deviation, if that is known.

```{r}
n <- 1000
mu <- 3.5
sigma <- 4
v <- rnorm(n, mu, sigma)
s <- sample(v, size = 30, replace = FALSE)
m <- mean(s)
m
```

```{r}
sd <- sd(s)
sd
```

```{r}
sem <- sd(s)/sqrt(length(s))
sem
```

```{r}
lower <- m - qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci <- c(lower, upper)
ci
```
*remember gives random variables*

**Central Limit Theorem**. The key importance of the CLT for us is that it states that the distribution of averages (or sums or other summary statistics…) of iid (independent and identically distributed) random variables becomes normal as the sample size increases. It is this fact that allows us to have a good sense of the mean and distribution of average events in a population even though we only observe one set of events and do not know what actual population distribution is. In fact, the CLT says nothing about the probability distribution for events in the original population, and that is exactly where its usefulness lies… that original probability distribution can be normal, skewed, all kinds of odd!


We are going to take lots of averages of samples from a particular non-normal distribution and then look at the distribution of those averages. Imagine we have some event that occurs in a population according to some probability mass function like the Poisson where we know λ=14. Recall, then, that the expectations of μ and σ2 for the Poisson distribution are both=λ.

Now let’s imagine taking a bunch of samples of size 10 from this population. We will take 1000 random samples of this size, calculate the average of each sample, and plot a histogram of those averages… it will be close to normal, and the standard deviation of the those average - i.e., of the sampling distribution - should be roughly equal to the estimated standard error, the square root of (λ/n). [Recall that λ is the expected variance, so this is simply the square root of (expected variance / sample size)]

```{r}
lambda <- 14
n <- 10
pop_se <- sqrt(lambda/n)  # the estimated SE
pop_se
```

```{r}
x <- NULL
for (i in 1:1000) {
    x[i] <- mean(rpois(n = n, lambda = lambda))
}
hist(x, breaks = seq(from = lambda - 4 * sqrt(lambda)/sqrt(n), to = lambda +
    4 * sqrt(lambda)/sqrt(n), length.out = 20), probability = TRUE)
```

```{r}
sd <- sd(x)  # st dev of the sampling distribution
sd
```

```{r}
qqnorm(x)
qqline(x)
```

Lower to sample of 100, the sampling distribution will be lower.
```{r}
n <- 100
pop_se <- sqrt(lambda/n)  # the estimated SE
pop_se
```

```{r}
x <- NULL
for (i in 1:1000) {
    x[i] <- mean(rpois(n = n, lambda = lambda))
}
hist(x, breaks = seq(from = lambda - 4 * sqrt(lambda)/sqrt(n), to = lambda +
    4 * sqrt(lambda)/sqrt(n), length.out = 20), probability = TRUE)
```

```{r}
sd <- sd(x)  # st dev of the sampling distribution
sd
```

```{r}
qqnorm(x)
qqline(x)
```
We can convert these distributions to standard normals by subtracting off the expected population mean (λ) and dividing by the standard error of the mean (an estimate of the standard deviation of the sampling distribution) and then plotting a histogram of those values along with a normal curve.
```{r}
curve(dnorm(x, 0, 1), -4, 4, ylim = c(0, 0.8))
z <- (x - lambda)/pop_se
hist(z, breaks = seq(from = -4, to = 4, length.out = 20), probability = TRUE,
    add = TRUE)
```

```{r}
n <- 100
x <- NULL
for (i in 1:1000) {
    x[i] <- sum(rpois(n = n, lambda = lambda))
} #using sum instead of mean
hist(x, breaks = seq(min(x), max(x), length.out = 20), probability = TRUE)
```
**Take Home Points:**

[1] The CLT states that, regardless of the underlying distribution, the distribution of averages (or sums or standard deviations, etc…) based on a large number of independent, identically distributed variables:

will be approximately normal,
will be centered at the population mean, and – will have a standard deviation roughly equal to the standard error of the mean.
Additionally, it suggests that variables that are expected to be the sum of multiple independent processes (e.g., measurement errors) will also have distributions that are nearly normal.

[2] Taking the mean and adding and subtracting the relevant standard normal quantile × the standard error yields a confidence interval for the mean, which gets wider as the coverage increases and gets narrower with less variability or larger sample sizes.

[3] As sample size increases, the standard error of the mean decreases and the distribution becomes more and more normal (i.e., has less skew and kurtosis, which are higher order moments of the distribution).

The **expected sampling distribution** for the proportion of successes is approximately normal centered at π and its standard deviation is estimated by sqrt(π(1−π)/n), which is, essentially, the standard error of the mean: it is the square root of (the expected variance / sample size). As above for μ, if we do not already have a population estimate for π, we can estimate this from a sample as  = x/n

Note: this expectation based on an approximation of the normal holds true only if both n×π and n×(1−π) are greater than roughly 5, so we should check this when working with proportion data.

## Challenge 1
Suppose a polling group in the Massachusetts is interested in the proportion of voting-age citizens in their state that already know they will vote for Elizabeth Warren in the upcoming November 5, 2024 general elections (don’t forget to vote!). The group obtains a yes or no answer from 1000 suitable randomly selected individuals. Of these individuals, 856 say they know they’ll vote for Senator Warren. How would we characterize the mean and variability associated with this proportion?
```{r}
n <- 1000
x <- 856
phat <- x/n  # our estimate of pi
phat
```
Are n×π and n×(1−π) both > 5? Yes!
```{r}
n * phat
```
```{r}
n * (1 - phat)
```
```{r}
pop_se <- sqrt((phat) * (1 - phat)/n)
```

So, what is the 95% CI around our estimate of the proportion of people who already know how they will vote?

```{r}
curve(dnorm(x, mean = phat, sd = pop_se), phat - 4 * pop_se, phat + 4 * pop_se)
upper <- phat + qnorm(0.975) * pop_se
lower <- phat - qnorm(0.975) * pop_se
ci <- c(lower, upper)
polygon(cbind(c(ci[1], seq(from = ci[1], to = ci[2], length.out = 1000), ci[2]),
    c(0, dnorm(seq(from = ci[1], to = ci[2], length.out = 1000), mean = phat,
        sd = pop_se), 0)), border = "black", col = "gray")
abline(v = ci)
abline(h = 0)
```

Thus far, we have discussed creating a confidence interval based on the CLT and the normal distribution, and our intervals took the form:

mean ± Z (the quantile from the standard normal curve) × standard error of the mean

But, when the size of our sample is small (n < 30), instead of using the normal distribution to calculate our CIs, statisticians typically use a different distribution to generate the relevant quantiles to multiply the standard error by… the t distribution (a.k.a., Gosset’s t or Student’s t distribution).

The choice of the appropriate t distribution to use in any particular statistical test is based on the number of degrees of freedom (df), i.e., the number of individual components in the calculation of a given statistic (such as the standard deviation) that are “free to change”.

Confidence intervals based on the t distribution are of the form:

mean ± T (the quantile from the t distribution) × standard error of the mean

The only change from those based on the normal distribution is that we’ve replaced the Z quantile of the standard normal with a T quantile.

Recall that a standard normal distribution can be generated by normalizing our sample (subtracting the population mean from each observation and then dividing all of these differences by the population standard deviation)…

(mean(x)-μ)/σ

The code below plots a standard normal distribution and then t distributions with varying degrees of freedom, specified using the df= argument. As for other distributions, R implements density, cumulative probability, quantile, and random functions for the t distribution.
```{r}
mu <- 0
sigma <- 1
curve(dnorm(x, mu, 1), mu - 4 * sigma, mu + 4 * sigma, main = "Normal Curve=red\nStudent's t=blue",
    xlab = "x", ylab = "f(x)", col = "red", lwd = 3)
for (i in c(1, 2, 3, 4, 5, 10, 20, 100)) {
    curve(dt(x, df = i), mu - 4 * sigma, mu + 4 * sigma, main = "T Curve", xlab = "x",
        ylab = "f(x)", add = TRUE, col = "blue", lty = 5)
}
```

```{r}
n <- 1e+05
mu <- 3.5
sigma <- 4
x <- rnorm(n, mu, sigma)
sample_size <- 30
s <- sample(x, size = sample_size, replace = FALSE)
m <- mean(s)
m
```

```{r}
sd <- sd(s)
sd
```

```{r}
sem <- sd(s)/sqrt(length(s))
sem
```

```{r}
lower <- m - qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci_norm <- c(lower, upper)
ci_norm
```

Now, let’s look at the CIs calculated based using the t distribution for the same sample size. For sample size 30, the difference in the CIs is negligible.
```{r}
lower <- m - qt(1 - 0.05/2, df = sample_size - 1) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qt(1 - 0.05/2, df = sample_size - 1) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci_t <- c(lower, upper)
ci_t
```

Using a sample size of 5:
```{r}
sample_size <- 5
s <- sample(x, size = sample_size, replace = FALSE)
m <- mean(s)
m
```

```{r}
sd <- sd(s)
sd
```

```{r}
sem <- sd(s)/sqrt(length(s))
sem
```

```{r}
lower <- m - qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci_norm <- c(lower, upper)
ci_norm
```

```{r}
lower <- m - qt(1 - 0.05/2, df = sample_size - 1) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qt(1 - 0.05/2, df = sample_size - 1) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci_t <- c(lower, upper)
ci_t
```

**End of Module 9**